The Form of Uncertainty Affects Selection for Social Learning

### 2023-04-04 ms.tex
removed edit tags

Social learner $i$ adopts their chosen 
teacher $j$'s observed average payoffs for each behavior, $b$, i.e., 
$\ledger \leftarrow \bar\pi_{jb}$; 
and agent $i$'s count of behavioral observations, $c_{ib}$, is set to 1 if teacher $j$ had at least
one observation of behavior $b$, and $c_{ib}$ is set to 0 otherwise. This way, expected payoffs
remain within $[0, 1]$, but social learners are flexible to adjust to environmental change. Setting $c_{ib}$ to 1 reflects the fact that the social learner treats the teacher's information as a single observation with the associated uncertainty, \edit{and serves as a useful baseline for understanding the interaction of different sources of uncertainty that we focus on here. In the real world this could represent the fact that some amount of information is always lost in social learning, through instruction or otherwise.  If a learner's $c_{ib}$ is set instead to the teacher's $c_{jb}$, or some sizable fraction of it, learners will more slowly update their mean payoff values, making individual learning less adaptive, which could be desirable if there is good reason to have more confidence in a teacher's information, or there are other biological or cultural factors that make social learning more trustworthy.} Asocial learners ($s_i = 0$) are initialized with
$c_{ib} = 0$ and $\ledger = 0$ for each behavior, $b$. % (Figure~\ref{fig:schematic}C). 
The newly initialized population then repeats the within-generation dynamics.

@@

When payoff ambiguity and selection-set size were small ($\pilow=0.1$ and $B=2$), and lifespan
was long ($L \geq 4$), asocial learning fixated \edit{because} periods of protracted
environmental instability \edit{caused social learners to more severely underperform compared
to asocial learners}, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs (Figure~\ref{fig:payoffTimeseries}A). Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. In other words, environmental change lowers social learners'
mean payoffs, but social-learner payoffs build back up as the information quality in the
population improves (blue dots, Figure~\ref{fig:payoffTimeseries}A). Even though a homogenous
population of social learners would outperform individual learners in this case, unpredictable
environmental stochasticity undermines the reliability of social learning. When lifespan was
minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite,
unexpected outcome could occur where social learning fixates despite the fact that a homogenous
population of asocial learners would outperform a homogenous population of social learners
(Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners \edit{since social learners greatly outperform asocial learners during longer periods of environmental stability}.
\begin{figure}
  \caption{Average number of generations (y-axes) to fixation. 
    Note the y-axis scale varies between plots, indicating different overall 
    time to fixation for different $(\pilow, B)$ combinations.} 
  \label{fig:steps}
\centering
    \includegraphics[width=\textwidth]{Figures/stepResultsPlots.pdf}
\end{figure}

>>>>

Social learner $i$ adopts their chosen 
teacher $j$'s observed average payoffs for each behavior, $b$, i.e., 
$\ledger \leftarrow \bar\pi_{jb}$; 
and agent $i$'s count of behavioral observations, $c_{ib}$, is set to 1 if teacher $j$ had at least
one observation of behavior $b$, and $c_{ib}$ is set to 0 otherwise. This way, expected payoffs
remain within $[0, 1]$, but social learners are flexible to adjust to environmental change. Setting $c_{ib}$ to 1 reflects the fact that the social learner treats the teacher's information as a single observation with the associated uncertainty, and serves as a useful baseline for understanding the interaction of different sources of uncertainty that we focus on here. In the real world this could represent the fact that some amount of information is always lost in social learning, through instruction or otherwise.  If a learner's $c_{ib}$ is set instead to the teacher's $c_{jb}$, or some sizable fraction of it, learners will more slowly update their mean payoff values, making individual learning less adaptive, which could be desirable if there is good reason to have more confidence in a teacher's information, or there are other biological or cultural factors that make social learning more trustworthy. Asocial learners ($s_i = 0$) are initialized with
$c_{ib} = 0$ and $\ledger = 0$ for each behavior, $b$. % (Figure~\ref{fig:schematic}C). 
The newly initialized population then repeats the within-generation dynamics.

@@

When payoff ambiguity and selection-set size were small ($\pilow=0.1$ and $B=2$), and lifespan
was long ($L \geq 4$), asocial learning fixated because periods of protracted
environmental instability caused social learners to more severely underperform compared
to asocial learners, even though expected
homogeneous social payoffs were greater than homogeneous asocial payoffs (Figure~\ref{fig:payoffTimeseries}A). Runs of environmental change strongly affected the geometric mean fitness of social learners who were weighing
outdated information too heavily. In other words, environmental change lowers social learners'
mean payoffs, but social-learner payoffs build back up as the information quality in the
population improves (blue dots, Figure~\ref{fig:payoffTimeseries}A). Even though a homogenous
population of social learners would outperform individual learners in this case, unpredictable
environmental stochasticity undermines the reliability of social learning. When lifespan was
minimal ($L=1$) and environmental variability was moderate ($u\approx0.5$), the opposite,
unexpected outcome could occur where social learning fixates despite the fact that a homogenous
population of asocial learners would outperform a homogenous population of social learners
(Figure~\ref{fig:payoffTimeseries}B). In this case, it takes relatively few consecutive generations with environmental stability to attract the population to all become social learners since social learners greatly outperform asocial learners during longer periods of environmental stability.
\begin{figure}
  \caption{Average number of generations (y-axes) to fixation. 
    Note the y-axis scale varies between plots, indicating different overall 
    time to fixation for different $(\pilow, B)$ combinations.} 
  \label{fig:steps}
\centering
    \includegraphics[width=\textwidth]{Figures/stepResultsPlots.pdf}
\end{figure}
